---
title: "Getting_and_Cleaning_Data_Course_Project"
author: "ajf"
date: "8/19/2022"
output: html_document
---

## Getting and Cleaning Data Course Project

In this project the goal is to put together in a single and tidy dataframe a collection of multiple dataframes following the information provided by the data collection README file

```{r Before kicking off}
# Set working directory as required so it contains the folder with the data 
setwd(".")
# load required libraries
library("dplyr")
```

## Data info
==================================================================
Human Activity Recognition Using Smartphones Dataset
Version 1.0
==================================================================
Jorge L. Reyes-Ortiz, Davide Anguita, Alessandro Ghio, Luca Oneto.
Smartlab - Non Linear Complex Systems Laboratory
DITEN - Universit‡ degli Studi di Genova.
Via Opera Pia 11A, I-16145, Genoa, Italy.
activityrecognition@smartlab.ws
www.smartlab.ws
==================================================================

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

For each record it is provided:
======================================

- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

The dataset includes the following files:
=========================================

- 'README.txt'

- 'features_info.txt': Shows information about the variables used on the feature vector.

- 'features.txt': List of all features.

- 'activity_labels.txt': Links the class labels with their activity name.

- 'train/X_train.txt': Training set.

- 'train/y_train.txt': Training labels.

- 'test/X_test.txt': Test set.

- 'test/y_test.txt': Test labels.

The following files are available for the train and test data. Their descriptions are equivalent. 

- 'train/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 

- 'train/Inertial Signals/total_acc_x_train.txt': The acceleration signal from the smartphone accelerometer X axis in standard gravity units 'g'. Every row shows a 128 element vector. The same description applies for the 'total_acc_x_train.txt' and 'total_acc_z_train.txt' files for the Y and Z axis. 

- 'train/Inertial Signals/body_acc_x_train.txt': The body acceleration signal obtained by subtracting the gravity from the total acceleration. 

- 'train/Inertial Signals/body_gyro_x_train.txt': The angular velocity vector measured by the gyroscope for each window sample. The units are radians/second. 

Notes: 
======
- Features are normalized and bounded within [-1,1].
- Each feature vector is a row on the text file.

For more information about this dataset contact: activityrecognition@smartlab.ws

License:
========
Use of this dataset in publications must be acknowledged by referencing the following publication [1] 

[1] Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012

This dataset is distributed AS-IS and no responsibility implied or explicit can be addressed to the authors or their institutions for its use or misuse. Any commercial use is prohibited.

Jorge L. Reyes-Ortiz, Alessandro Ghio, Luca Oneto, Davide Anguita. November 2012.


## Codes

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

tBodyAcc-XYZ
tGravityAcc-XYZ
tBodyAccJerk-XYZ
tBodyGyro-XYZ
tBodyGyroJerk-XYZ
tBodyAccMag
tGravityAccMag
tBodyAccJerkMag
tBodyGyroMag
tBodyGyroJerkMag
fBodyAcc-XYZ
fBodyAccJerk-XYZ
fBodyGyro-XYZ
fBodyAccMag
fBodyAccJerkMag
fBodyGyroMag
fBodyGyroJerkMag

The set of variables that were estimated from these signals are: 

mean(): Mean value
std(): Standard deviation
mad(): Median absolute deviation 
max(): Largest value in array
min(): Smallest value in array
sma(): Signal magnitude area
energy(): Energy measure. Sum of the squares divided by the number of values. 
iqr(): Interquartile range 
entropy(): Signal entropy
arCoeff(): Autorregresion coefficients with Burg order equal to 4
correlation(): correlation coefficient between two signals
maxInds(): index of the frequency component with largest magnitude
meanFreq(): Weighted average of the frequency components to obtain a mean frequency
skewness(): skewness of the frequency domain signal 
kurtosis(): kurtosis of the frequency domain signal 
bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.
angle(): Angle between to vectors.

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

gravityMean
tBodyAccMean
tBodyAccJerkMean
tBodyGyroMean
tBodyGyroJerkMean

The complete list of variables of each feature vector is available in 'features.txt'


## Data gathering

It is necessary to read all files provided and assign names to the variables.
Let's start with the general files.

The features files contains two variables describing the variables in x_test/train files.
However, variables are not unique. I apply function make.unique to solve this. Now, features$unique can be used as colnames vector.

The activity_labels file contains the code for the activities. Colnames assign to identify the variables with the labels and activities corresponding to them.

```{r General files}
# Read features
features <- read.table("./UCI HAR Dataset/features.txt")
features$unique <- make.unique(as.character(features$V2),sep="_")
# Read activity labels
activity_labels <- read.table("./UCI HAR Dataset/activity_labels.txt")
colnames(activity_labels) <- c("labels","activity")
```

Then, data is divided into two groups: train and test.
Let's organize them separately first. 

## Reading and assigning variable names to training data

Dataframes are obtained from txt files.
Variables are names according to content.
x_train varibles are rename with features in a descriptive manner.
For Internal signal data 9 files are read and variables named using the file name plus a measurement number.

All this files are binded since each row correspond to the same feature vector

```{r Training data}
# Read training data
x_train <- read.table("./UCI HAR Dataset/train/X_train.txt")
colnames(x_train) <- features$V2
y_train <- read.table("./UCI HAR Dataset/train/y_train.txt")
colnames(y_train) <- c("labels")
subject_train <- read.table("./UCI HAR Dataset/train/subject_train.txt")
colnames(subject_train) <- c("subject")

body_acc_x_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_acc_x_train.txt")
body_acc_y_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_acc_y_train.txt")
body_acc_z_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_acc_z_train.txt")

body_gyro_x_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_gyro_x_train.txt")
body_gyro_y_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_gyro_y_train.txt")
body_gyro_z_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_gyro_z_train.txt")

total_acc_x_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/total_acc_x_train.txt")
total_acc_y_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/total_acc_y_train.txt")
total_acc_z_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/total_acc_z_train.txt")

colnames(body_acc_x_train) <- paste("body_acc_x_",c(1:length(colnames(body_acc_x_train))),sep="")
colnames(body_acc_y_train) <- paste("body_acc_y_",c(1:length(colnames(body_acc_y_train))),sep="")
colnames(body_acc_z_train) <- paste("body_acc_z_",c(1:length(colnames(body_acc_z_train))),sep="")

colnames(body_gyro_x_train) <- paste("body_gyro_x_",c(1:length(colnames(body_gyro_x_train))),sep="")
colnames(body_gyro_y_train) <- paste("body_gyro_y_",c(1:length(colnames(body_gyro_y_train))),sep="")
colnames(body_gyro_z_train) <- paste("body_gyro_z_",c(1:length(colnames(body_gyro_z_train))),sep="")

colnames(total_acc_x_train) <- paste("total_acc_x_",c(1:length(colnames(total_acc_x_train))),sep="")
colnames(total_acc_y_train) <- paste("total_acc_y_",c(1:length(colnames(total_acc_y_train))),sep="")
colnames(total_acc_z_train) <- paste("total_acc_z_",c(1:length(colnames(total_acc_z_train))),sep="")


train <- cbind(subject_train ,y_train, x_train, 
               body_acc_x_train, body_acc_y_train,body_acc_z_train,
               body_gyro_x_train, body_gyro_y_train,body_gyro_z_train,
               total_acc_x_train, total_acc_y_train,total_acc_z_train)
```


## Reading and assigning variable names to testing data

The same is applied to test files

```{r Testing data}
# Read test data
# Appropriately labels the data set with descriptive variable names.

x_test <- read.table("./UCI HAR Dataset/test/X_test.txt")
colnames(x_test) <- features$V2
y_test <- read.table("./UCI HAR Dataset/test/Y_test.txt")
colnames(y_test) <- c("labels")
subject_test <- read.table("./UCI HAR Dataset/test/subject_test.txt")
colnames(subject_test) <- c("subject")

body_acc_x_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_acc_x_test.txt")
body_acc_y_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_acc_y_test.txt")
body_acc_z_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_acc_z_test.txt")

body_gyro_x_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_gyro_x_test.txt")
body_gyro_y_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_gyro_y_test.txt")
body_gyro_z_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_gyro_z_test.txt")

total_acc_x_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/total_acc_x_test.txt")
total_acc_y_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/total_acc_y_test.txt")
total_acc_z_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/total_acc_z_test.txt")

colnames(body_acc_x_test) <- paste("body_acc_x_",c(1:length(colnames(body_acc_x_test))),sep="")
colnames(body_acc_y_test) <- paste("body_acc_y_",c(1:length(colnames(body_acc_y_test))),sep="")
colnames(body_acc_z_test) <- paste("body_acc_z_",c(1:length(colnames(body_acc_z_test))),sep="")

colnames(body_gyro_x_test) <- paste("body_gyro_x_",c(1:length(colnames(body_gyro_x_test))),sep="")
colnames(body_gyro_y_test) <- paste("body_gyro_y_",c(1:length(colnames(body_gyro_y_test))),sep="")
colnames(body_gyro_z_test) <- paste("body_gyro_z_",c(1:length(colnames(body_gyro_z_test))),sep="")

colnames(total_acc_x_test) <- paste("total_acc_x_",c(1:length(colnames(total_acc_x_test))),sep="")
colnames(total_acc_y_test) <- paste("total_acc_y_",c(1:length(colnames(total_acc_y_test))),sep="")
colnames(total_acc_z_test) <- paste("total_acc_z_",c(1:length(colnames(total_acc_z_test))),sep="")


test <- cbind(subject_test ,y_test, x_test, 
               body_acc_x_test, body_acc_y_test,body_acc_z_test,
               body_gyro_x_test, body_gyro_y_test,body_gyro_z_test,
               total_acc_x_test, total_acc_y_test,total_acc_z_test)
```

## Combining Training and Testing data

Since the structure of the data is equivalent I cbind both dataframes
To finish tidying the data I will label the activities with descriptive variables too by merging the dataframe that codes the activity labels with the one containing the measurements.

```{r cbind}
# combine test and train data
all_data <- rbind(train,test)

# Uses descriptive activity names to name the activities in the data set
merge_activity <- merge(activity_labels, all_data, by="labels")
````

## Extracting info

Here two options to get the mean and the sd for each variable.

# Option 1
Generate loop to get the mean and the sd from the columns of the datafrane with all the data.
Store the info in two separate vectors and put them together in a new dataframe using colnames to identify the measurement in each row.

```{r Option 1: mean and sd}
# Extracts only the measurements on the mean and standard deviation for each measurement. 
all_mean <- NULL
all_sd <- NULL
for(i in 3:length(colnames(all_data))){
  all_mean[i] <- mean(all_data[,i])
  all_sd[i] <- sd(all_data[,i])
}
measures <- data.frame(colnames(all_data),all_mean, all_sd)
```

# Option 2
Turn the variables levels and subject into factors since they are not measuring anything. They are codes.
Use the function summarise_if to get the average/sd of all numeric variables.

```{r Option 2: mean and sd}
merge_activity$subject <- as.factor(merge_activity$subject)
merge_activity$labels <- as.factor(merge_activity$labels)
mean_all <- merge_activity %>% summarise_if(is.numeric, mean, na.rm = TRUE)
sd_all <- merge_activity %>% summarise_if(is.numeric, sd, na.rm = TRUE)
```

# Calculate means for groups
Group as needed and summarise using the function of interes - here mean.
In average_all we store means by subjects and activities.
In average_subject, activities are not taken into account and stores the mean for each subject.
In average_acivity, subjects are not taken into account and stores the mean for each activity.

```{r Groups}
# From the data set in step 4, creates a second, independent tidy data set 
# with the average of each variable for each activity and each subject.

merge_activity <- group_by(merge_activity, activity, subject)
average_all <- summarize_all(merge_activity, mean)

average_activity <- merge_activity %>% group_by(activity) %>% 
  summarise_if(is.numeric, mean, na.rm = TRUE)

average_subject <- merge_activity %>% group_by(subject) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)
````

## Saving results
The dataframe that contains all the info is saved into a file called tidy_data.csv

```{r Write file}
# generate a file with the whole set of tidy data
write.csv(merge_activity, file="tidy_data.csv")
aux <- read.csv("tidy_data.csv", header=TRUE)
        
````
