---
title: "Getting_and_Cleaning_Data_Course_Project"
author: "ajf"
date: "8/19/2022"
output: html_document
---

## Getting and Cleaning Data Course Project

In this project the goal is to put together in a single and tidy dataframe a collection of multiple dataframes following the information provided by the data collection README file

```{r Before kicking off}
# Set working directory as required so it contains the folder with the data 
setwd(".")
# load required libraries
library("dplyr")
```

## Data gathering

It is necessary to read all files provided and assign names to the variables.
Let's start with the general files.

The features files contains two variables describing the variables in x_test/train files.
However, variables are not unique. I apply function make.unique to solve this. Now, features$unique can be used as colnames vector.

The activity_labels file contains the code for the activities. Colnames assign to identify the variables with the labels and activities corresponding to them.

```{r General files}
# Read features
features <- read.table("./UCI HAR Dataset/features.txt")
features$unique <- make.unique(as.character(features$V2),sep="_")
# Read activity labels
activity_labels <- read.table("./UCI HAR Dataset/activity_labels.txt")
colnames(activity_labels) <- c("labels","activity")
```

Then, data is divided into two groups: train and test.
Let's organize them separately first. 

## Reading and assigning variable names to training data

Dataframes are obtained from txt files.
Variables are names according to content.
x_train varibles are rename with features in a descriptive manner.
For Internal signal data 9 files are read and variables named using the file name plus a measurement number.

All this files are binded since each row correspond to the same feature vector

```{r Training data}
# Read training data
x_train <- read.table("./UCI HAR Dataset/train/X_train.txt")
colnames(x_train) <- features$V2
y_train <- read.table("./UCI HAR Dataset/train/y_train.txt")
colnames(y_train) <- c("labels")
subject_train <- read.table("./UCI HAR Dataset/train/subject_train.txt")
colnames(subject_train) <- c("subject")

body_acc_x_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_acc_x_train.txt")
body_acc_y_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_acc_y_train.txt")
body_acc_z_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_acc_z_train.txt")

body_gyro_x_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_gyro_x_train.txt")
body_gyro_y_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_gyro_y_train.txt")
body_gyro_z_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/body_gyro_z_train.txt")

total_acc_x_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/total_acc_x_train.txt")
total_acc_y_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/total_acc_y_train.txt")
total_acc_z_train <- read.table("./UCI HAR Dataset/train/Inertial Signals/total_acc_z_train.txt")

colnames(body_acc_x_train) <- paste("body_acc_x_",c(1:length(colnames(body_acc_x_train))),sep="")
colnames(body_acc_y_train) <- paste("body_acc_y_",c(1:length(colnames(body_acc_y_train))),sep="")
colnames(body_acc_z_train) <- paste("body_acc_z_",c(1:length(colnames(body_acc_z_train))),sep="")

colnames(body_gyro_x_train) <- paste("body_gyro_x_",c(1:length(colnames(body_gyro_x_train))),sep="")
colnames(body_gyro_y_train) <- paste("body_gyro_y_",c(1:length(colnames(body_gyro_y_train))),sep="")
colnames(body_gyro_z_train) <- paste("body_gyro_z_",c(1:length(colnames(body_gyro_z_train))),sep="")

colnames(total_acc_x_train) <- paste("total_acc_x_",c(1:length(colnames(total_acc_x_train))),sep="")
colnames(total_acc_y_train) <- paste("total_acc_y_",c(1:length(colnames(total_acc_y_train))),sep="")
colnames(total_acc_z_train) <- paste("total_acc_z_",c(1:length(colnames(total_acc_z_train))),sep="")


train <- cbind(subject_train ,y_train, x_train, 
               body_acc_x_train, body_acc_y_train,body_acc_z_train,
               body_gyro_x_train, body_gyro_y_train,body_gyro_z_train,
               total_acc_x_train, total_acc_y_train,total_acc_z_train)
```


## Reading and assigning variable names to testing data

The same is applied to test files

```{r Testing data}
# Read test data
# Appropriately labels the data set with descriptive variable names.

x_test <- read.table("./UCI HAR Dataset/test/X_test.txt")
colnames(x_test) <- features$V2
y_test <- read.table("./UCI HAR Dataset/test/Y_test.txt")
colnames(y_test) <- c("labels")
subject_test <- read.table("./UCI HAR Dataset/test/subject_test.txt")
colnames(subject_test) <- c("subject")

body_acc_x_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_acc_x_test.txt")
body_acc_y_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_acc_y_test.txt")
body_acc_z_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_acc_z_test.txt")

body_gyro_x_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_gyro_x_test.txt")
body_gyro_y_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_gyro_y_test.txt")
body_gyro_z_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/body_gyro_z_test.txt")

total_acc_x_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/total_acc_x_test.txt")
total_acc_y_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/total_acc_y_test.txt")
total_acc_z_test <- read.table("./UCI HAR Dataset/test/Inertial Signals/total_acc_z_test.txt")

colnames(body_acc_x_test) <- paste("body_acc_x_",c(1:length(colnames(body_acc_x_test))),sep="")
colnames(body_acc_y_test) <- paste("body_acc_y_",c(1:length(colnames(body_acc_y_test))),sep="")
colnames(body_acc_z_test) <- paste("body_acc_z_",c(1:length(colnames(body_acc_z_test))),sep="")

colnames(body_gyro_x_test) <- paste("body_gyro_x_",c(1:length(colnames(body_gyro_x_test))),sep="")
colnames(body_gyro_y_test) <- paste("body_gyro_y_",c(1:length(colnames(body_gyro_y_test))),sep="")
colnames(body_gyro_z_test) <- paste("body_gyro_z_",c(1:length(colnames(body_gyro_z_test))),sep="")

colnames(total_acc_x_test) <- paste("total_acc_x_",c(1:length(colnames(total_acc_x_test))),sep="")
colnames(total_acc_y_test) <- paste("total_acc_y_",c(1:length(colnames(total_acc_y_test))),sep="")
colnames(total_acc_z_test) <- paste("total_acc_z_",c(1:length(colnames(total_acc_z_test))),sep="")


test <- cbind(subject_test ,y_test, x_test, 
               body_acc_x_test, body_acc_y_test,body_acc_z_test,
               body_gyro_x_test, body_gyro_y_test,body_gyro_z_test,
               total_acc_x_test, total_acc_y_test,total_acc_z_test)
```

## Combining Training and Testing data

Since the structure of the data is equivalent I cbind both dataframes
To finish tidying the data I will label the activities with descriptive variables too by merging the dataframe that codes the activity labels with the one containing the measurements.

```{r cbind}
# combine test and train data
all_data <- rbind(train,test)

# Uses descriptive activity names to name the activities in the data set
merge_activity <- merge(activity_labels, all_data, by="labels")
````

## Extracting info

Here two options to get the mean and the sd for each variable.

# Option 1
Generate loop to get the mean and the sd from the columns of the datafrane with all the data.
Store the info in two separate vectors and put them together in a new dataframe using colnames to identify the measurement in each row.

```{r Option 1: mean and sd}
# Extracts only the measurements on the mean and standard deviation for each measurement. 
all_mean <- NULL
all_sd <- NULL
for(i in 3:length(colnames(all_data))){
  all_mean[i] <- mean(all_data[,i])
  all_sd[i] <- sd(all_data[,i])
}
measures <- data.frame(colnames(all_data),all_mean, all_sd)
```

# Option 2
Turn the variables levels and subject into factors since they are not measuring anything. They are codes.
Use the function summarise_if to get the average/sd of all numeric variables.

```{r Option 2: mean and sd}
merge_activity$subject <- as.factor(merge_activity$subject)
merge_activity$labels <- as.factor(merge_activity$labels)
mean_all <- merge_activity %>% summarise_if(is.numeric, mean, na.rm = TRUE)
sd_all <- merge_activity %>% summarise_if(is.numeric, sd, na.rm = TRUE)
```

# Calculate means for groups
Group as needed and summarise using the function of interes - here mean.
In average_all we store means by subjects and activities.
In average_subject, activities are not taken into account and stores the mean for each subject.
In average_acivity, subjects are not taken into account and stores the mean for each activity.

```{r Groups}
# From the data set in step 4, creates a second, independent tidy data set 
# with the average of each variable for each activity and each subject.

merge_activity <- group_by(merge_activity, activity, subject)
average_all <- summarize_all(merge_activity, mean)

average_activity <- merge_activity %>% group_by(activity) %>% 
  summarise_if(is.numeric, mean, na.rm = TRUE)

average_subject <- merge_activity %>% group_by(subject) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)
````

## Saving results
The dataframe that contains all the info is saved into a file called tidy_data.csv

```{r Write file}
# generate a file with the whole set of tidy data
write.csv(merge_activity, file="tidy_data.csv")
aux <- read.csv("tidy_data.csv", header=TRUE)
        
````
