---
title: "Course Project:Developing Data Products"
author: "ajf"
date: '2022-11-12'
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Course Project

This App shows an histogram with the frecuency of cars with a determined miles per galon performance using mtcars dataframe and selecting a range of cylinder numbers, horsepower and weigth of the cars.

Select ranges using the sliders.

Results will be shown on the right.

Variables:  
- Number of cylinders  
- Gross horsepower  
- Weight (1000 lbs)  

Output:  
- Miles/(US) gallon Histogram

## Slide with ui.R code

```{r, echo = TRUE, eval=FALSE}
library(shiny)

# Define UI for application that draws a histogram
shinyUI(fluidPage(

    # Application title
    titlePanel("Three aspects of automobile design and performance"),

    # Sidebar with a slider input for number of ...
    sidebarLayout(
        sidebarPanel(
            sliderInput("cyl",
                        "Number of cylinders:",
                        min = 4,
                        max = 8,
                        value = -c(5,7)),
            sliderInput("hp",
                        "Gross horsepower:",
                        min = 52,
                        max = 335,
                        value = c(100,193)),
            sliderInput("wt",
                        "Weight (1000 lbs):",
                        min = 1.513,
                        max = 5.424,
                        value = c(2,3.47)),
        ),

        # Show a plot of the generated distribution
        mainPanel(
            h3("Miles/(US) gallon:"),
            plotOutput("plot")
        )
    )
))
```

## Slide with server.R code

```{r, echo = TRUE, eval=FALSE}
library(shiny)
library(dplyr)

# Define server logic required to draw a histogram
shinyServer(function(input, output) {
  output$plot <- renderPlot({
    min_cyl <- input$cyl[1]
    max_cyl <- input$cyl[2]
    min_hp <- input$hp[1]
    max_hp <- input$hp[2]
    min_wt <- input$wt[1]
    max_wt <- input$wt[2]
    
    x <- filter(mtcars, mtcars$cyl>=min_cyl) 
    x <- filter(x, x$cyl<=max_cyl) 
    x <- filter(x, x$hp>=min_hp) 
    x <- filter(x, x$hp<=max_hp) 
    x <- filter(x, x$wt>=min_wt) 
    x <- filter(x, x$wt<=max_wt)
    mpg <- x[,1]
    #  cyl <- seq(min(x), max(x), length.out = input$cyl + 1)
    
    # draw the histogram with the specified number of bins
    hist(mpg, breaks = 10, col = 'darkgray', border = 'white',
         xlab = 'mpg',
         main = 'Histogram of Miles/(US) gallon')
    
  })
  
})
```

## Example of Output Histogram

```{r, echo = TRUE, eval=TRUE}
  hist(mtcars$mpg, breaks = 10, col = 'darkgray', border = 'white',
         xlab = 'mpg',
         main = 'Histogram of Miles/(US) gallon')
```
